---
title: "IPAT 2 — Antiparasitic Chemical Discharge in Norwegian Salmon Aquaculture (2012–2019)"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

```{r libraries}
library(tidyverse)
library(scales)
library(patchwork)
```

---

## 1. Load data

We load the IPAT 2 dataset covering Norwegian salmon aquaculture from 2012 to 2019.
The impact variable **A\_out** (kg/year) is the annual mass of persistent antiparasitic
substances discharged into Norwegian fjords, defined as:

$$A_{out} = A_{bath,excl.H_2O_2} \times 1.0 \;\;+\;\; A_{oral} \times 0.03$$

H2O2 is excluded because it degrades in under one hour in seawater and would otherwise
represent over 99.9% of total mass every year, completely masking the ecologically
relevant signal. The 3% factor for oral treatments reflects that approximately 97% of
orally administered substances are metabolised by the fish before reaching the fjord.

```{r load}
df <- read_csv("ipat2_antiparasitic_discharge.csv")

# Compute IPAT factors
df <- df %>%
  mutate(
    N_lice_total = N_salmon * N_lice_per_fish,        # lice/year
    lambda       = N_lice_per_fish,                    # lice/fish
    tau          = T_total / N_lice_total,             # events/lice
    phi          = T_chem_excl_h2o2 / T_total,        # dimensionless
    delta        = A_out_kg / T_chem_excl_h2o2,       # kg/event
    # Index base 2012 = 100 for decoupling chart
    idx_P        = P_tonnes / first(P_tonnes) * 100,
    idx_A_out    = A_out_kg / first(A_out_kg) * 100
  )

glimpse(df)
```

---

## 2. Absolute decoupling — A\_out vs production

The central result of the IPAT 2 analysis: persistent chemical discharge fell **-95.3%**
while salmon production grew **+10.7%** between 2012 and 2019. Both series are indexed
to 2012 = 100 so they can be compared on the same scale despite having different units
(tonnes vs kg). When the two lines diverge in opposite directions, this confirms
**strong absolute decoupling** — the industry grew while its chemical footprint shrank.

```{r decoupling, fig.height=5}
df_long <- df %>%
  select(year, idx_P, idx_A_out) %>%
  pivot_longer(-year, names_to = "series", values_to = "index") %>%
  mutate(series = recode(series,
    idx_P     = "Salmon production (P_tonnes, index 2012=100)",
    idx_A_out = "A_out — persistent discharge (index 2012=100)"
  ))

ggplot(df_long, aes(x = year, y = index, colour = series, linetype = series)) +
  geom_hline(yintercept = 100, colour = "grey80", linewidth = 0.4) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_colour_manual(values = c(
    "Salmon production (P_tonnes, index 2012=100)"      = "#1D9E75",
    "A_out — persistent discharge (index 2012=100)" = "#D85A30"
  )) +
  scale_linetype_manual(values = c(
    "Salmon production (P_tonnes, index 2012=100)"      = "solid",
    "A_out — persistent discharge (index 2012=100)" = "dashed"
  )) +
  scale_x_continuous(breaks = 2012:2019) +
  scale_y_continuous(labels = label_number(suffix = "")) +
  labs(
    title    = "Absolute decoupling: persistent chemical discharge vs salmon production",
    subtitle = "Index 2012 = 100  |  A_out = A_bath_excl_H2O2 x 1.0 + A_oral x 0.03",
    x = NULL, y = "Index (2012 = 100)",
    colour = NULL, linetype = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())

ggsave("Images/Absolut_Decoupling_IPAT_2.png", width = 9, height = 5, dpi = 300)

```

---

## 3. A\_out in absolute terms (kg per year)

This chart shows the raw magnitude of persistent chemical discharge each year. The sharp
drop from 5 186 kg in 2014 to 293 kg in 2017 corresponds to the industry-wide shift away
from bath treatments (100% discharge) toward oral treatments (3% discharge) and
non-chemical alternatives, following regulatory pressure and the development of cleaner
fish and mechanical delousing technologies.

```{r aout_abs, fig.height=4.5}
ggplot(df, aes(x = year, y = A_out_kg)) +
  geom_col(fill = "#D85A30", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = comma(round(A_out_kg))),
            vjust = -0.4, size = 3.5, colour = "#D85A30") +
  scale_x_continuous(breaks = 2012:2019) +
  scale_y_continuous(labels = label_comma(),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "A_out — persistent antiparasitic substances discharged into Norwegian fjords",
    subtitle = "kg per year  |  excludes H2O2 (half-life < 1 h)",
    x = NULL, y = "kg per year"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor    = element_blank(),
        panel.grid.major.x  = element_blank())

ggsave("Images/A_out_absolut_term.png", width = 9, height = 4.5, dpi = 300)

```

---

## 4. IPAT factor time series

The IPAT 2 identity decomposes A\_out into five multiplicative factors:

$$A_{out} = N_{salmon} \;\times\; \frac{N_{lice}}{N_{salmon}} \;\times\; \frac{Tr_{total}}{N_{lice}} \;\times\; \frac{Tr_{chem,pol}}{Tr_{total}} \;\times\; \frac{A_{out}}{Tr_{chem,pol}}$$

Each panel below tracks one factor independently over time, making it possible to
isolate which levers moved and when. A downward trend reduces discharge; an upward
trend increases it. The panels should be read alongside the waterfall chart (section 5)
which quantifies each factor's contribution to the total change.

```{r ipat_factors, fig.height=9, fig.width=10}
p1 <- ggplot(df, aes(year, P_tonnes / 1e6)) +
  geom_line(colour = "#1D9E75", linewidth = 1) +
  geom_point(colour = "#1D9E75", size = 2.5) +
  scale_x_continuous(breaks = 2012:2019) +
  labs(title = "P — Production scale",
       y = "Million tonnes per year", x = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p2 <- ggplot(df, aes(year, lambda)) +
  geom_line(colour = "#BA7517", linewidth = 1) +
  geom_point(colour = "#BA7517", size = 2.5) +
  scale_x_continuous(breaks = 2012:2019) +
  labs(title = "lambda — Lice pressure  (N_lice / N_salmon)",
       y = "lice per fish", x = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p3 <- ggplot(df, aes(year, tau * 1e4)) +
  geom_line(colour = "#378ADD", linewidth = 1) +
  geom_point(colour = "#378ADD", size = 2.5) +
  scale_x_continuous(breaks = 2012:2019) +
  labs(title = "tau — Treatment intensity  (Tr_total / N_lice)",
       y = "events per lice  (x10^-4)", x = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p4 <- ggplot(df, aes(year, phi)) +
  geom_line(colour = "#7F77DD", linewidth = 1) +
  geom_point(colour = "#7F77DD", size = 2.5) +
  scale_x_continuous(breaks = 2012:2019) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "phi — Chemical treatment share  (Tr_chem,pol / Tr_total)",
       y = "dimensionless", x = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p5 <- ggplot(df, aes(year, delta)) +
  geom_line(colour = "#D85A30", linewidth = 1) +
  geom_point(colour = "#D85A30", size = 2.5) +
  scale_x_continuous(breaks = 2012:2019) +
  labs(title = "delta — Discharge per chemical event  (A_out / Tr_chem,pol)",
       y = "kg per event", x = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p6 <- ggplot() + theme_void()

combined_plot <- (p1 + p2) / (p3 + p4) / (p5 + p6) +
  plot_annotation(
    title    = "IPAT 5-factor decomposition — individual factor trends (2012–2019)",
    subtitle = "A_out = N_salmon x lambda x tau x phi x delta",
    theme    = theme(
      plot.title    = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 11, colour = "grey40")
    )
  )

combined_plot
ggsave("Images/IPAT_2_Factors.png", plot = combined_plot, width = 10, height = 9, dpi = 300)
```

---

## 5. IPAT waterfall — ln-change contributions (2012 to 2019)

This chart quantifies each factor's contribution to the total change in A\_out between
2012 and 2019, expressed as log-changes so that contributions are additive. The five
driver bars sum exactly to the total bar (A\_out = -3.05 ln-points = -95.3%).
**Orange bars** push discharge upward; **green bars** drive it down. The dominant driver
is **delta** (-2.04), which captures the bath-to-oral substitution: switching from bath
treatments (100% discharge fraction) to oral treatments (3% discharge fraction) reduced
the kg released per treatment event by 87%, independently of how many treatments were
performed or how many lice were present.

```{r waterfall, fig.height=4.5}
ipat <- tibble(
  factor = factor(
    c("N_salmon\n(scale)", "lambda\n(lice/fish)", "tau\n(treat/lice)",
      "phi\n(chem share)", "delta\n(kg/event)", "A_out\n(total)"),
    levels = c("N_salmon\n(scale)", "lambda\n(lice/fish)", "tau\n(treat/lice)",
               "phi\n(chem share)", "delta\n(kg/event)", "A_out\n(total)")
  ),
  dln = c(
    log(last(df$N_salmon)  / first(df$N_salmon)),
    log(last(df$lambda)    / first(df$lambda)),
    log(last(df$tau)       / first(df$tau)),
    log(last(df$phi)       / first(df$phi)),
    log(last(df$delta)     / first(df$delta)),
    log(last(df$A_out_kg)  / first(df$A_out_kg))
  ),
  type = c("driver", "driver", "driver", "driver", "driver", "total")
)

ipat <- ipat %>%
  mutate(fill_col = case_when(
    type == "total" ~ "#2C2C2A",
    dln  >  0       ~ "#D85A30",
    TRUE            ~ "#1D9E75"
  ))

ggplot(ipat, aes(x = factor, y = dln, fill = fill_col)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_text(aes(label = sprintf("%+.2f", dln),
                vjust = ifelse(dln >= 0, -0.4, 1.3)),
            size = 3.5, colour = "grey30") +
  scale_fill_identity() +
  scale_y_continuous(breaks = seq(-2.5, 0.5, 0.5)) +
  labs(
    title    = "IPAT decomposition — ln-change contributions (2012 to 2019)",
    subtitle = "Total delta-ln(A_out) = -3.05  |  Green = reduces discharge, Orange = increases discharge",
    x = NULL, y = "Delta-ln"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor    = element_blank(),
        panel.grid.major.x  = element_blank())
ggsave("Images/IPAT_2_Waterfall.png", width = 8.5, height = 4.5, dpi = 300, create.dir = TRUE)
```

---

## 6. Treatment mix — stacked bar (events per year)

This chart shows the structural shift in intervention types over time. It explains the
movement of two IPAT factors simultaneously: **phi** (chemical share of treatments,
driven by the rise of non-chemical and cleaner fish alternatives) and **delta** (kg per
chemical event, driven by the shift from bath to oral within chemical treatments). The
near-disappearance of bath persistent treatments after 2016 is the single most important
structural change in the dataset. Note the cleaner fish collapse in 2018-2019, which
reflects a reporting reclassification following the 2017-2018 welfare crisis rather than
a complete cessation of use.

```{r treatment_mix, fig.height=5}
treat <- df %>%
  select(year, T_bath_excl_h2o2, T_chem_h2o2, T_oral,
         T_non_chem, T_cleaner_fish) %>%
  pivot_longer(-year, names_to = "type", values_to = "events") %>%
  mutate(type = factor(type, levels = c(
    "T_bath_excl_h2o2", "T_chem_h2o2", "T_oral",
    "T_non_chem", "T_cleaner_fish"
  )))

type_labels <- c(
  T_bath_excl_h2o2 = "Bath — persistent (azamethiphos, pyrethroids)",
  T_chem_h2o2      = "Bath — H2O2",
  T_oral           = "Oral — persistent (emamectin, benzoylphenyl ureas)",
  T_non_chem       = "Non-chemical (mechanical / thermal)",
  T_cleaner_fish   = "Cleaner fish  [reporting collapse 2018-2019]"
)

type_colours <- c(
  T_bath_excl_h2o2 = "#D85A30",
  T_chem_h2o2      = "#378ADD",
  T_oral           = "#639922",
  T_non_chem       = "#888780",
  T_cleaner_fish   = "#7F77DD"
)

ggplot(treat, aes(x = year, y = events, fill = type)) +
  geom_col(width = 0.7, alpha = 0.9) +
  scale_fill_manual(values = type_colours, labels = type_labels) +
  scale_x_continuous(breaks = 2012:2019) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Treatment mix — all intervention types (events per year)",
    subtitle = "Shift from bath persistent to oral + non-chemical drives phi and delta reductions",
    x = NULL, y = "Treatment events per year", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position    = "bottom",
        legend.text        = element_text(size = 9),
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_blank()) +
  guides(fill = guide_legend(ncol = 1))
ggsave("Images/Treatment_Mix_shift_IPAT_2.png", width = 9, height = 5, dpi = 300)

```

---

## 7. Identity verification

As a methodological check, we verify that the product of the five IPAT factors
exactly equals the raw A\_out\_kg value for every year. A residual of 0% confirms
that the decomposition is algebraically exact and that no approximation has been
introduced in the factor construction.

```{r identity_check}
df %>%
  mutate(
    ipat_product = N_salmon * lambda * tau * phi * delta,
    residual_pct = (ipat_product - A_out_kg) / A_out_kg * 100
  ) %>%
  select(year, A_out_kg, ipat_product, residual_pct) %>%
  mutate(across(c(A_out_kg, ipat_product), ~round(., 2)),
         residual_pct = round(residual_pct, 6)) %>%
  knitr::kable(
    col.names = c("Year", "A_out_kg (raw)", "IPAT product", "Residual (%)"),
    align     = "rrrr",
    caption   = "IPAT identity verification — residual should be 0% all years"
  )
```
