---
title: "IPAT 1 — Wild Fish Extraction for Norwegian Salmon Feed"

date: "Spring 2026"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Period: 2000–2013 | Species: Atlantic salmon only

## Identity: W = P × (F/P) × (M/F) × (W/M)

## Step 0 — Load Libraries

```{r libraries}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(scales)

theme_set(theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "grey80")
  ))
```

---

## Step 1 — Load P (Atlantic Salmon Sales from Fiskeridir)

**Key fix**: The `Totalt/Total` row contains Excel `=SUM()` formulas which R cannot evaluate.
We instead **sum the county rows directly** to obtain the national total.

```{r load_P}
extract_atlantic_salmon_totals <- function(filepath) {
  sheets <- excel_sheets(filepath)
  results <- list()
  seen_years <- c()

  for (sheet_name in sheets) {
    df <- read_excel(filepath, sheet = sheet_name, col_names = FALSE)
    nrows <- nrow(df)

    for (start_row in 1:(nrows - 5)) {
      year_row <- suppressWarnings(as.numeric(as.character(unlist(df[start_row, ]))))
      years_found <- year_row[!is.na(year_row) & year_row >= 1990 & year_row <= 2030]
      if (length(years_found) < 2) next

      year_cols <- which(!is.na(year_row) & year_row >= 1990 & year_row <= 2030)

      # Find the Total row index
      total_row_idx <- NA
      for (j in (start_row + 1):min(start_row + 30, nrows)) {
        cell <- as.character(df[j, 1])
        if (grepl("Totalt|Total", cell, ignore.case = TRUE)) {
          total_row_idx <- j
          break
        }
      }
      if (is.na(total_row_idx)) next

      # Data rows = skip 2 col-name rows after header, up to (not including) Total row
      data_start <- start_row + 3
      data_end   <- total_row_idx - 1
      if (data_end < data_start) next

      # For each year, sum Laks column (at year_col index) across county rows
      for (col_idx in year_cols) {
        yr <- as.integer(year_row[col_idx])
        if (yr %in% seen_years) next

        county_vals <- suppressWarnings(
          as.numeric(as.character(unlist(df[data_start:data_end, col_idx])))
        )
        total_laks <- sum(county_vals, na.rm = TRUE)

        if (!is.na(total_laks) && total_laks > 10000) {
          results[[length(results) + 1]] <- data.frame(
            year = yr,
            atlantic_salmon_t = total_laks
          )
          seen_years <- c(seen_years, yr)
        }
      }
      break
    }
  }

  bind_rows(results) |> arrange(year)
}

SALG_FILE <- "data_sales.xlsx"
df_raw <- extract_atlantic_salmon_totals(SALG_FILE)

df_P <- df_raw |>
  filter(year >= 2000, year <= 2013) |>
  arrange(year)

kable(df_P, caption = "P — Atlantic salmon sales (tonnes round weight), 2000–2013",
      format.args = list(big.mark = ","))
```

### Plot P

```{r plot_P, fig.width=9, fig.height=4}
ggplot(df_P, aes(x = year, y = atlantic_salmon_t / 1e6)) +
  geom_area(alpha = 0.15, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 1.2) +
  geom_point(colour = "steelblue", size = 3) +
  geom_text(
    data = df_P[c(1, nrow(df_P)), ],
    aes(label = paste0(round(atlantic_salmon_t / 1e6, 2), " Mt")),
    vjust = -1, size = 3.2, colour = "steelblue"
  ) +
  scale_x_continuous(breaks = df_P$year) +
  labs(
    title = "P — Atlantic salmon production in Norway (2000–2013)",
    subtitle = "Source: Fiskeridir data_sales.xlsx",
    x = "Year", y = "Million tonnes (round weight)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Images/Atlantic_Salmon_production.png", width = 9, height = 4, dpi = 300)


```

---

## Step 2 — Nofima Benchmark Data + Linear Interpolation

```{r nofima_benchmarks}
df_nofima <- data.frame(
  year = c(2000,  2010,  2012,  2013),
  M_F  = c(0.646, 0.414, 0.307, 0.292),
  FCR  = c(1.50,  1.20,  1.15,  1.15),
  W_M  = c(1.80,  1.10,  0.70,  0.70)
)
kable(df_nofima, caption = "Nofima benchmark data")
```

### Linear Interpolation to Annual Values

```{r interpolation}
all_years <- df_P$year

df_interp <- data.frame(
  year = all_years,
  M_F  = approx(df_nofima$year, df_nofima$M_F, xout = all_years)$y,
  FCR  = approx(df_nofima$year, df_nofima$FCR, xout = all_years)$y,
  W_M  = approx(df_nofima$year, df_nofima$W_M, xout = all_years)$y
)

kable(round(df_interp, 4), caption = "Interpolated annual values")
```

### Plot Interpolated Series vs Benchmark Points

```{r plot_interpolated, fig.width=14, fig.height=4}
plot_factor <- function(var, title, ylabel, color) {
  ggplot() +
    geom_line(data = df_interp, aes(x = year, y = .data[[var]]),
              colour = color, linewidth = 1.2) +
    geom_point(data = df_nofima, aes(x = year, y = .data[[var]]),
               colour = color, size = 3, shape = 21,
               fill = "white", stroke = 1.5) +
    geom_text(data = df_nofima,
              aes(x = year, y = .data[[var]],
                  label = round(.data[[var]], 3)),
              vjust = -1.2, size = 2.8, colour = color) +
    scale_x_continuous(breaks = all_years) +
    labs(title = title, x = "Year", y = ylabel) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

p1 <- plot_factor("M_F", "M/F — Marine ingredient share",
                  "Fraction of feed (kg/kg)", "#008080")
p2 <- plot_factor("FCR", "F/P — Feed conversion ratio",
                  "kg feed / kg salmon", "darkorange")
p3 <- plot_factor("W_M", "W/M — FFDR_FM (forage fish)",
                  "kg wild fish / kg marine", "firebrick")

combined_interp <- p1 + p2 + p3 +
  plot_annotation(
    title = "Piecewise linear interpolation between Nofima benchmarks (2000–2013)",
    subtitle = "Sources: Ytrestøyl et al. (2015) + Nofima Report 36/2014"
  )

combined_interp
ggsave("Images/Ipat1_Interpolation_factors.png", plot = combined_interp, width = 14, height = 4, dpi = 300, create.dir = TRUE)
```

---

## Step 3 — Compute W

```{r compute_W}
df <- df_P |>
  left_join(df_interp, by = "year") |>
  mutate(
    F_t = atlantic_salmon_t * FCR,
    M_t = F_t * M_F,
    W_t = M_t * W_M
  )

df |>
  select(year,
         `P (t)` = atlantic_salmon_t,
         `F/P`   = FCR,
         `M/F`   = M_F,
         `W/M`   = W_M,
         `F (t)` = F_t,
         `M (t)` = M_t,
         `W (t)` = W_t) |>
  mutate(
    across(c(`P (t)`, `F (t)`, `M (t)`, `W (t)`), round),
    across(c(`F/P`, `M/F`, `W/M`), ~round(., 3))
  ) |>
  kable(caption = "IPAT decomposition — Atlantic salmon, 2000–2013",
        format.args = list(big.mark = ","))
```

### Plot W alongside P — Decoupling

```{r plot_W_vs_P, fig.width=10, fig.height=4}
df |>
  select(year, P = atlantic_salmon_t, W = W_t) |>
  pivot_longer(-year, names_to = "variable", values_to = "tonnes") |>
  ggplot(aes(x = year, y = tonnes / 1e6, colour = variable, fill = variable)) +
  geom_area(alpha = 0.12, position = "identity") +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("P" = "steelblue", "W" = "firebrick"),
                      labels = c("P = salmon produced", "W = wild fish extracted")) +
  scale_fill_manual(values = c("P" = "steelblue", "W" = "firebrick"), guide = "none") +
  scale_x_continuous(breaks = df$year) +
  labs(title = "W vs P — Absolute decoupling of wild fish extraction from salmon production",
       subtitle = "P grew +166% while W fell -64% over 2000-2013",
       x = "Year", y = "Million tonnes", colour = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

ggsave("Images/Absolut_decoupling.png", width = 10, height = 4, dpi = 300)

```

---

## Step 4 — Growth Factor Analysis (2000 to 2013)

Ln-additive decomposition: ln(W_t/W_0) = ln(P_t/P_0) + ln(FCR_t/FCR_0) + ln(MF_t/MF_0) + ln(WM_t/WM_0)

```{r growth_factors}
base_yr <- df |> filter(year == 2000)
end_yr  <- df |> filter(year == 2013)

delta_lnP   <- log(end_yr$atlantic_salmon_t / base_yr$atlantic_salmon_t)
delta_lnFCR <- log(end_yr$FCR / base_yr$FCR)
delta_lnMF  <- log(end_yr$M_F / base_yr$M_F)
delta_lnWM  <- log(end_yr$W_M / base_yr$W_M)
delta_lnW   <- log(end_yr$W_t / base_yr$W_t)

gf <- data.frame(
  Factor     = c("P (production)", "F/P (eFCR)", "M/F (marine share)", "W/M (FFDR_FM)", "W (total, observed)"),
  Value_2000 = round(c(base_yr$atlantic_salmon_t, base_yr$FCR, base_yr$M_F, base_yr$W_M, base_yr$W_t)),
  Value_2013 = round(c(end_yr$atlantic_salmon_t,  end_yr$FCR,  end_yr$M_F,  end_yr$W_M,  end_yr$W_t)),
  Pct_change = round(c(
    (end_yr$atlantic_salmon_t/base_yr$atlantic_salmon_t - 1)*100,
    (end_yr$FCR/base_yr$FCR - 1)*100,
    (end_yr$M_F/base_yr$M_F - 1)*100,
    (end_yr$W_M/base_yr$W_M - 1)*100,
    (end_yr$W_t/base_yr$W_t - 1)*100
  ), 1),
  Ln_contrib = round(c(delta_lnP, delta_lnFCR, delta_lnMF, delta_lnWM, delta_lnW), 3),
  Share_pct  = round(c(delta_lnP, delta_lnFCR, delta_lnMF, delta_lnWM, delta_lnW) /
                       delta_lnW * 100, 1)
)

kable(gf,
      caption = "Ln-additive growth factor decomposition: 2000 to 2013",
      col.names = c("Factor", "2000", "2013", "% change", "ln contribution", "Share of Delta ln(W) (%)"))
```

### Waterfall chart

```{r plot_waterfall, fig.width=9, fig.height=5}
wf_data <- data.frame(
  label   = c("P\n(+166%)", "F/P\n(-23%)", "M/F\n(-55%)", "W/M\n(-61%)", "W\n(result)"),
  contrib = c(delta_lnP, delta_lnFCR, delta_lnMF, delta_lnWM, delta_lnW),
  type    = c("driver", "efficiency", "efficiency", "efficiency", "result")
)

# Compute bar positions
wf_data$xpos  <- seq_len(nrow(wf_data))
wf_data$yend  <- cumsum(c(0, wf_data$contrib[-nrow(wf_data)]))
wf_data$yend[nrow(wf_data)] <- 0
wf_data$ystart <- wf_data$yend
wf_data$ytop   <- wf_data$yend + wf_data$contrib
wf_data$ystart[nrow(wf_data)] <- 0
wf_data$ytop[nrow(wf_data)]   <- wf_data$contrib[nrow(wf_data)]

ggplot(wf_data) +
  geom_rect(aes(xmin = xpos - 0.4, xmax = xpos + 0.4,
                ymin = pmin(ystart, ytop), ymax = pmax(ystart, ytop),
                fill = type), alpha = 0.85) +
  geom_hline(yintercept = 0, colour = "grey30", linewidth = 0.5) +
  geom_text(aes(x = xpos, y = (ystart + ytop) / 2,
              label = sprintf("%+.2f", contrib)),
          size = 3.8, fontface = "bold", colour = "white") +
  scale_x_continuous(breaks = wf_data$xpos, labels = wf_data$label) +
  scale_fill_manual(
    values = c("driver" = "steelblue", "efficiency" = "#008080", "result" = "firebrick"),
    labels = c("driver" = "Growth driver (pushes W up)",
               "efficiency" = "Efficiency gain (pushes W down)",
               "result" = "Net result on W"),
    name = NULL) +
  labs(title = "Waterfall decomposition of Delta ln(W) (2000 to 2013)",
       subtitle = "Production growth was more than offset by three simultaneous efficiency gains",
       x = NULL, y = "ln contribution to Delta ln(W)") +
  theme(legend.position = "top")

ggsave("Images/Waterfall_chart_factors_IPAT_1.png", width = 9, height = 5, dpi = 300)

```

### Index plot (2000 = 100)

```{r plot_index, fig.width=10, fig.height=5}
df_index <- df |>
  mutate(
    idx_P   = atlantic_salmon_t / first(atlantic_salmon_t) * 100,
    idx_FCR = FCR / first(FCR) * 100,
    idx_MF  = M_F / first(M_F) * 100,
    idx_WM  = W_M / first(W_M) * 100,
    idx_W   = W_t / first(W_t) * 100
  ) |>
  select(year,
         `P (salmon)`        = idx_P,
         `F/P (eFCR)`        = idx_FCR,
         `M/F (marine share)`= idx_MF,
         `W/M (FFDR_FM)`     = idx_WM,
         `W (wild fish)`     = idx_W) |>
  pivot_longer(-year, names_to = "variable", values_to = "index")

ggplot(df_index, aes(x = year, y = index, colour = variable,
                     linewidth = (variable == "W (wild fish)"))) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey50") +
  geom_line() +
  geom_point(size = 2) +
  scale_colour_manual(values = c(
    "P (salmon)"         = "steelblue",
    "F/P (eFCR)"         = "darkorange",
    "M/F (marine share)" = "#008080",
    "W/M (FFDR_FM)"      = "firebrick",
    "W (wild fish)"      = "black"
  )) +
  scale_linewidth_manual(values = c("FALSE" = 0.9, "TRUE" = 2), guide = "none") +
  scale_x_continuous(breaks = df$year) +
  labs(title = "Index of all IPAT factors (2000 = 100)",
       subtitle = "P grew +166%; efficiency gains more than compensated — W fell to 36% of 2000 level",
       x = "Year", y = "Index (2000 = 100)", colour = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

ggsave("Images/Index_IPAT_1.png", width = 10, height = 5, dpi = 300)

```

---

## Step 5 — Contribution Charts

### 5a — Multiplicative factor contributions relative to 2000

```{r contribution_multipliers, fig.width=10, fig.height=5}
df_contrib <- df |>
  mutate(
    contrib_P   = atlantic_salmon_t / first(atlantic_salmon_t),
    contrib_FCR = FCR / first(FCR),
    contrib_MF  = M_F / first(M_F),
    contrib_WM  = W_M / first(W_M)
  ) |>
  select(year,
         `P (production)` = contrib_P,
         `F/P (eFCR)`     = contrib_FCR,
         `M/F (marine)`   = contrib_MF,
         `W/M (FFDR_FM)`  = contrib_WM) |>
  pivot_longer(-year, names_to = "factor", values_to = "multiplier")

ggplot(df_contrib, aes(x = year, y = multiplier, colour = factor)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(
    "P (production)" = "steelblue",
    "F/P (eFCR)"     = "darkorange",
    "M/F (marine)"   = "#008080",
    "W/M (FFDR_FM)"  = "firebrick"
  )) +
  scale_x_continuous(breaks = df$year) +
  labs(title = "Multiplicative factor contributions relative to 2000",
       subtitle = "Values above 1 increase W; below 1 decrease W. Their product = W index.",
       x = "Year", y = "Factor ratio (2000 = 1)", colour = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

ggsave("Images/Contribution_multipliers_IPAT_1.png", width = 10, height = 5, dpi = 300)

```

### 5b — Year-on-year change in W decomposed by factor

```{r yoy_decomposition, fig.width=12, fig.height=5}
df_yoy <- df |>
  mutate(
    dW_P   = W_t * log(atlantic_salmon_t / lag(atlantic_salmon_t)),
    dW_FCR = W_t * log(FCR / lag(FCR)),
    dW_MF  = W_t * log(M_F / lag(M_F)),
    dW_WM  = W_t * log(W_M / lag(W_M))
  ) |>
  filter(!is.na(dW_P)) |>
  select(year,
         `P (production)` = dW_P,
         `F/P (eFCR)`     = dW_FCR,
         `M/F (marine)`   = dW_MF,
         `W/M (FFDR_FM)`  = dW_WM) |>
  pivot_longer(-year, names_to = "factor", values_to = "delta_W")

ggplot(df_yoy, aes(x = year, y = delta_W / 1e3, fill = factor)) +
  geom_col(position = "stack", alpha = 0.85) +
  geom_hline(yintercept = 0, colour = "grey30") +
  scale_fill_manual(values = c(
    "P (production)" = "steelblue",
    "F/P (eFCR)"     = "darkorange",
    "M/F (marine)"   = "#008080",
    "W/M (FFDR_FM)"  = "firebrick"
  )) +
  scale_x_continuous(breaks = unique(df_yoy$year)) +
  labs(title = "Year-on-year change in W decomposed by IPAT factor",
       subtitle = "Bars above zero push W up (P); bars below push W down (efficiency factors)",
       x = "Year", y = "Delta W (thousand tonnes)", fill = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

ggsave("Images/YoY_decomposition_IPAT_1.png", width = 12, height = 5, dpi = 300)

```

---

## Step 6 — Export to CSV

```{r export_csv}
csv_out <- df |>
  select(
    year,
    P_atlantic_salmon_t    = atlantic_salmon_t,
    FP_eFCR                = FCR,
    MF_marine_share        = M_F,
    WM_FFDR_FM             = W_M,
    F_total_feed_t         = F_t,
    M_marine_ingredients_t = M_t,
    W_wild_fish_t          = W_t
  )

write.csv(csv_out, "ipat_decomposition_2000_2013.csv", row.names = FALSE)
cat("Saved: ipat_decomposition_2000_2013.csv\n")

csv_out |>
  mutate(
    across(ends_with("_t"), round),
    across(c(FP_eFCR, MF_marine_share, WM_FFDR_FM), ~round(., 3))
  ) |>
  kable(caption = "Final IPAT decomposition table",
        format.args = list(big.mark = ","))
```
